---
title: "Hurricane wind field model"
output: html_document
editor_options: 
  chunk_output_type: console
---


## IBTraCS data

https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/shapefile/

```{r}
L <- "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/shapefile/IBTrACS.NA.list.v04r00.lines.zip"
  
if(!"IBTrACS.NA.list.v04r00.lines.zip" %in% list.files(here::here("data"))) {
download.file(url = L,
              destfile = here::here("data",
                                    "IBTrACS.NA.list.v04r00.lines.zip"))
unzip(here::here("data", "IBTrACS.NA.list.v04r00.lines.zip"),
      exdir = here::here("data"))
}

Tracks.sf <- sf::st_read(dsn = here::here("data"), 
                         layer = "IBTrACS.NA.list.v04r00.lines") |>
  sf::st_transform(crs = 32616)
```

Find all storms that are within 100 km of a location
```{r}
point <- sf::st_point(c(-85.3, 29.9))
location.sf <- sf::st_sfc(point, 
                           crs = 4326) |>
  sf::st_transform(crs = 32616) |>
  sf::st_as_sf()

buffer.sf <- location.sf |>
  sf::st_buffer(dist = 100000)

Tracks2.sf <- Tracks.sf |>
  sf::st_intersection(buffer.sf)

Tracks3.sf <- Tracks2.sf[!is.na(Tracks2.sf$USA_ATCFID),]

length(unique(Tracks3.sf$USA_ATCFID))
max(Tracks3.sf$SEASON) - min(Tracks3.sf$SEASON) + 1
```

List the unique storms by `USA_ATCFID` = `track_id`
```{r}
unique(Tracks3.sf$USA_ATCFID)
```

Group by storm ID and find the fastest wind
```{r}
X <- Tracks3.sf |>
  sf::st_drop_geometry() |>
  dplyr::group_by(USA_ATCFID) |>
  dplyr::summarize(Wmax = max(WMO_WIND, na.rm = TRUE))
```

```{r}
tmap::tmap_mode("view")
tmap::tm_shape(Tracks2.sf) +
  tmap::tm_lines()
```

## Wind field model

<https://github.com/jbcannon/hurrecon/tree/v.0.0.1>

```{r}
install.packages('devtools')
devtools::install_github('jbcannon/hurrecon')

library(hurrecon)
path <- here::here("data", "hurdat_data.csv")
fetch_best_tracks_data(path,
                       src = "https://www.nhc.noaa.gov/data/hurdat/hurdat2-1851-2022-040723.txt")
```

The last 6 numbers before the `.txt` are the date of the archive. As of 16 May 2023, the archive date is 7 April 2023

Load the data for Hurricane Michael (2018, ID = AL142018) as a simple feature data frame with CRS UTM16 (EPSG 32616)
```{r}
Michael_Track.sf <- load_hurdat_track(path, 
                                      trackID = 'AL142018')
X.sf <- load_hurdat_track(path,
                          trackID = 'AL041851')
```

Get land mask
```{r}
library(terra)
data("geographic")
```

Create raster of winds with units of m/s. This takes a long time (13 minutes on desktop)
```{r}
t0 <- Sys.time()
Wind.r <- hurrecon_run(Michael_Track.sf, 
                       max_rad_km = 100, 
                       res_m = 30000, 
                       max_interp_dist_km = 5)
Sys.time() - t0
```

Make a plot
```{r}
plot(land)
plot(Wind.r, add = TRUE)
plot(land, add = TRUE)
```

Extract the wind speed from the raster at a particular location.

Create the location
```{r}
point <- sf::st_point(c(-85.3, 29.9))
location.sf <- sf::st_sfc(point, 
                           crs = 4326) |>
  sf::st_transform(crs = sf::st_crs(Wind.r)) |>
  sf::st_as_sf()

Wind.r |>
  terra::extract(location.sf)
```


Include only track observations when the track is within 200 km of a location
```{r}
buffer.sf <- location.sf |>
  sf::st_buffer(dist = 200000)

Michael_Track2.sf <- Michael_Track.sf |>
  sf::st_intersection(buffer.sf) 

colnames(Michael_Track2.sf)[11:22] <- colnames(Michael_Track.sf)[11:22]
```

Run the wind field model
```{r}
t0 <- Sys.time()
Wind.r2 <- hurrecon_run(Michael_Track2.sf, 
                        max_rad_km = 100, res_m = 5000, max_interp_dist_km = 5)
Sys.time() - t0
```

32 seconds

```{r}
tmap::tmap_mode("view")
tmap::tm_shape(Wind.r2) +
  tmap::tm_raster(alpha = .5) +
  tmap::tm_shape(Michael_Track.sf) +
    tmap::tm_dots()
```

Multiple storms wind fields for a specific location into a single `SpatVector`.

Get all tracks
```{r}
df <- read.csv(here::here("data", "hurdat_data.csv"))
df$date <- lubridate::ymd(df$date)
df$lon2 <- as.numeric(df$lon)
df$lat2 <- as.numeric(df$lat)
df <- df[complete.cases(df[, c("lon2")]), ]

sdf <- df |>
  sf::st_as_sf(coords = c("lon2", "lat2"))
sf::st_crs(sdf) <- 4326

sdf <- sdf |>
  sf::st_transform(crs = 32616)

sdf[, c("utmx", "utmy")] <- sf::st_coordinates(sdf)

cn <- c("34kt_ne", "34kt_se",	"34kt_sw", "34kt_nw",	
        "50kt_ne", "50kt_se",	"50kt_sw", "50kt_nw",
        "64kt_ne", "64kt_se",	"64kt_sw", "64kt_nw")
colnames(sdf)[11:22] <- cn
```

Find all storms that have passed with 